Supporting information: Tail associations and their impact on extinction risk

### 2019-11-06 SuppMat.Rmd
deleting all \nameref{} link
added a derivation of the Spearman for extreme tail dependent noise

<!--# Algorithm to generate extreme tail association  \label{extr_tail_dep}-->
Bivarite environmental noise with extreme left-tail association (as in Fig. 
\ref{MT-fig_intro_cop}A) was generated using an algorithm originally presented 
by @Ghosh_copula (their Appendix S10), presented again here for convenience.
If $T$ time steps of noise were desired, step 1 was to generate a $T$ by $2$
matrix, $M$, of independent random numbers, each standard-normally distributed. 
Then, each row of $M$ was identified for modification randomly and
independently, with a 50\% chance. For those rows identified, every element
of the row was replaced by the negative of the absolute value of the first element
of the row, including the first element itself. For the other rows, every 
element in the row was replaced by its own absolute value. For noise
with extreme right-tail association (as in Fig. 
\ref{MT-fig_intro_cop}D), take the negative of the entire matrix. For
both types of noise, every entry was transformed by the cumulative distribution
function of a standard normal distribution, to make each column uniformly
distributed.

<!--Spearman correlation of the extreme-tail-associated noise-->
The Spearman correlation of the extreme left- or right-tail associated 
noise was `r srho_common`, up to sampling variation.
<!--***DAN: Dan to fill this in with a derivation of this fact-->

@@

>>>>

<!--# Algorithm to generate extreme tail association  \label{extr_tail_dep}-->
Bivarite environmental noise with extreme left-tail association (as in Fig. 
\ref{MT-fig_intro_cop}A) was generated using an algorithm originally presented 
by @Ghosh_copula (their Appendix S10), presented again here for convenience.
If $T$ time steps of noise were desired, step 1 was to generate a $T$ by $2$
matrix, $M$, of independent random numbers, each standard-normally distributed. 
Then, each row of $M$ was identified for modification randomly and
independently, with a 50\% chance. For those rows identified, every element
of the row was replaced by the negative of the absolute value of the first element
of the row, including the first element itself. For the other rows, every 
element in the row was replaced by its own absolute value. For noise
with extreme right-tail association (as in Fig. 
\ref{MT-fig_intro_cop}D), take the negative of the entire matrix. For
both types of noise, every entry was transformed by the cumulative distribution
function of a standard normal distribution, to make each column uniformly
distributed.
The Spearman correlation of the extreme left- or right-tail associated 
noise was `r srho_common`, up to sampling variation. A derivation
of this fact is presented below.

@@

<!--derivation of the fact that Spearman is 0.875 for the extreme tail dependent noise-->
We now present a derivation of the fact that, up to sampling variance,
the Spearman correlation of our extreme tail associated noise was 
$0.875$. Let $u_i$, $v_i$ be the ranks of a sample of noise, $i=1,\ldots,N$.
Assuming that $N$ is divisible by $4$ makes the presentation of the 
derivation simpler, so we adopt that assumption. We show the derivation for the left-tail
associated noise, the right-tail case being similar. Then 
\vspace{-10pt}
\begin{equation}
\rho=\frac{\sum_{i=1}^N (u_i - \mean(u_i))(v_i-\mean(v_i))}{(N-1)\sqrt{\var(u_i)\var(v_i)}}.
\end{equation}
\noindent However, it is straightforward to see that $\mean(u_i) = \mean(v_i)=(N+1)/2$,
which we henceforth call $\mu$. It is also straightforward to see
that $\var(u_i)=\var(v_i)=\frac{\sum_{i=1}^N (i-\mu)^2}{N-1}$, since 
the set consisting of the $u_i$ is the same as the set consisting of the 
$v_i$, and is also the same as the set consisting of the numbers $1,\ldots,N$. 
Therefore,
\vspace{-10pt}
\begin{equation}
\rho=\frac{\sum_{i=1}^N (u_i-\mu)(v_i-\mu)}{\sum_{i=1}^N (i-\mu)^2}.
\end{equation}
\noindent We can reindex the $u_i$ and $v_i$ so that $u_i=v_i=i$ for $i=1,\ldots,N/2$.
This uses the fact that we are working with the left-tail associated noise.
Then we have 
\vspace{-20pt}
\begin{align}
\rho &= \frac{\sum_{i=1}^{N/2} (i-\mu)^2 +\sum_{i=N/2+1}^N (u_i-\mu)(v_i-\mu)}{\sum_{i=1}^N (i-\mu)^2} \\
&= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (u_i-\mu)(v_i-\mu)}{\sum_{i=1}^N (i-\mu)^2}.\label{eq:rho1}
\end{align}
\noindent We define $\mu_2 = \mean_{i=N/2+1}^N(u_i)=\mean_{i=N/2+1}^N(i)$. It is straightforward
to show that this is $3N/4+1/2$. Then the numerator of the right term in 
(\ref{eq:rho1}) can be rearranged as
\vspace{-20pt}
\begin{align}
\sum_{i=N/2+1}^N (u_i-\mu)(v_i-\mu) &= \sum_{i=N/2+1}^N (u_i-\mu_2+\mu_2-\mu)(v_i-\mu_2+\mu_2-\mu) \\
&= \sum_{i=N/2+1}^N \left( (u_i-\mu_2)(v_i-\mu_2)+(u_i-\mu_2)(\mu_2-\mu)
  +(v_i-\mu_2)(\mu_2-\mu)+(\mu_2-\mu)^2 \right) \\
&= \sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2) + \frac{N}{2}(\mu_2-\mu)^2 \\
&= \sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2) + \frac{N^3}{32}.
\end{align}
\noindent The denominator of the right term in (\ref{eq:rho1}) can be rearranged as
\vspace{-20pt}
\begin{align}
\sum_{i=1}^N (i-\mu)^2 &= \sum_{i=1}^N (i^2-2\mu i +\mu^2) \\
&= \sum_{i=1}^N i^2 - 2\mu\sum_{i=1}^N i +N\mu^2.
\end{align}
\noindent Applying Faulhaber's formula to simplify the first term, this is
\vspace{-20pt}
\begin{align}
\sum_{i=1}^N (i-\mu)^2 &= \frac{N(N+1)(2N+1)}{6}-2\mu\frac{N(N+1)}{2}+N\mu^2 \label{eq:rho2}\\
&= \frac{N(N+1)(N-1)}{12},\label{eq:rho3}
\end{align}
\noindent where we have omitted the simple algebra leading from (\ref{eq:rho2}) 
to (\ref{eq:rho3}). Plugging in our expressions for the numerator and denominator,
(\ref{eq:rho1}) becomes
\vspace{-20pt}
\begin{align}
\rho &= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (u_i-\mu)(v_i-\mu)}{\sum_{i=1}^N (i-\mu)^2} \\
&= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2) + \frac{N^3}{32}}{\frac{N(N+1)(N-1)}{12}} \\
&= \frac{1}{2}+\frac{12\sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2)}{N(N+1)(N-1)}+
\frac{3N^3}{8N(N+1)(N-1)} \label{eq:rho4}
\end{align}
\noindent The second term of (\ref{eq:rho4}) is proportional to the covariance 
$\cov_{i=N/2+1}^N (u_i,v_i)$. To see this, note that
\vspace{-20pt}
\begin{align}
\cov_{i=N/2+1}^N (u_i,v_i) &= \frac{\sum_{i=N/2+1}^N(u_i-\mu_2)(v_i-\mu_2)}{(N/2-1)
\frac{\sum_{i=N/2+1}^N (i-\mu_2)^2}{N/2-1}} \\
&= \frac{\sum_{i=N/2+1}^N(u_i-\mu_2)(v_i-\mu_2)}{\sum_{i=N/2+1}^N (i-\mu_2)^2}.\label{eq:rho5}
\end{align}
\noindent But it is straightforward to show that
\vspace{-10pt}
\begin{equation}
\sum_{i=N/2+1}^N (i-\mu_2)^2 = \frac{N(N+2)(N-2)}{96}.
\end{equation}
\noindent Plugging this into (\ref{eq:rho5}), we get
\vspace{-20pt}
\begin{align}
\cov_{i=N/2+1}^N (u_i,v_i) &= \frac{96\sum_{i=N/2+1}^N(u_i-\mu_2)(v_i-\mu_2)}{N(N+2)(N-2)}.
\end{align}
\noindent Therefore the middle term of (\ref{eq:rho4}) equals
\vspace{-10pt}
\begin{equation}
\frac{12\sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2)}{N(N+1)(N-1)} = \cov_{i=N/2+1}^N (u_i,v_i) \left( \frac{N(N+2)(N-2)}{8N(N+1)(N-1)} \right).
\end{equation}
\noindent The limit of this as $N \rightarrow \infty$ is $0$ because the covariance
goes to 0. Thus the limit of (\ref{eq:rho4}) as $N \rightarrow \infty$
is $7/8=0.875$, as originally claimed.
